---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Paul Brencick (A17668863)"
format: pdf
toc: TRUE
---

## Background

Pertussis (aka Whooping Cough) is a common lung infection caused by the bateria B. Pertussis. 

This can infect people of any age but it is most severe and deadly against infants. 

The CDC track the [number of reported cases in the US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html) 

We can "scrape" this data with the **datapasta** package!

```{r}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22530,
    21996
  )
)
```

> Q1. Make a plot of `year` vs `cases`

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis Cases per Year 1922-2023") +
  scale_y_continuous(labels = scales::label_number())
  
```

> Q. Add some major milestones including the first wP vaccine rollout (1946), the switch to the newer aP vaccine (1996), and the Covid years (2020).

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis Cases per Year 1922-2023") +
  scale_y_continuous(labels = scales::label_number()) +
  geom_vline(xintercept = 1946, color="red", linetype = "dashed") +
  geom_vline(xintercept = 1996,color = "blue", linetype = "dashed") +
  geom_vline(xintercept = 2020,color = "orange", linetype = "dashed") 
  
```

There we high case numbers in the pre-1940s as a result of no available vaccines. Following 1946, the release of the wP vaccine was widely distributed leading to a dramatic decrease in cases. In 1996, the new aP vaccine was release which didn't really come with a decrease in cases. We see the rise of cases once again in 2004 which may have been likely to the waning immunity caused from the vaccines. Additionally the anti-vax movement and political landscapes may have decreased the amount of people getting the vaccine. Another factor could be that our testing was getting better hence the increased amount of cases. In 2020, we see a dramatic decrease in cases due to the isolation and social distance guidelines put in place which prevented respiratory based illnesses. 

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## CMI-PB Project

[Computational Models of Immunity - Pertussis Boost](https://www.cmi-pb.org/) aims to provide the scientific community with this very information.

They make there data available via JSON format API. We an read this in R with the `read_json()` function from the **jsonlite** package

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", TRUE)

head(subject)
```

> Q. Hw many wP and aP individuals area in this `subject` table 

```{r}
table(subject$infancy_vac)
```
> Q. What is the biological sex breakdown?

```{r}
table(subject$biological_sex)
```

> Q. In terms of race and gender is this dataset representitive of the US population? 

```{r}
table(subject$race, subject$biological_sex)

```

This dataset is not representitve of the US population as it was taken during Covid.



Let's read some more database tables!

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", TRUE)

ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To analyze this data we need to first "join" (merge/link) the different tables so we have all the data in one place not spread across different tables.

We can use the `*_join()` famility of functions from the **dplyer** to do this. 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
```

```{r}
abdata <- inner_join(ab_titer, meta)

head(abdata, 5)
```

> Q. What Antibody isotypes are measured for these patients?

```{r}
table(abdata$isotype)
```

> Q. What antigens are reported?

```{r}
table(abdata$antigen)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
Most recent data set has a lot more data compared to the data sets from 2021-2022. However, it is significantly less than the 2020 data set.

Lets focus on IgG antigen and make a plot of MFI_normaized for all antigens. 

```{r}
igg <- abdata |>
  filter(isotype =="IgG")

head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)

```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q. Is there a difference of aP vs wP individuals with these values?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, color = infancy_vac) +
  geom_boxplot()
```
> Q. Is there a temprol response ie. do values increase or decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, color = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

## Focus on "PT" Pertusisis toxin antigen

```{r}
pt.igg21 <- igg |> filter(antigen =="PT",
              dataset == "2021_dataset")
```

```{r}
ggplot(pt.igg21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14, col="black",lty="dashed")

```

```{r}
ggplot(pt.igg21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.4) +
  geom_vline(xintercept = 14, col="black",lty="dashed") +
  geom_vline(xintercept = 0, col="black", lty="dashed") +
  geom_smooth(aes(group = infancy_vac), method = "loess", span = 1.06, se = FALSE, linewidth = 1.5) +
  
  labs(title = "CMI-PB 2021 dataset IgG PT",
       subtitle = "Dashed lines at day 0 (pre boost) and day 14 (post boost)",
       x = "Day relative to boost",
       y = "Normalised MFI")
```



