---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Paul Brencick (A17668863)"
format: pdf
toc: True
---

## Background

Today we will perform an RNA seq analysis on the effects of Dexamethasone (hereafter "Dex"), a common steroid, on airway smooth muscles (ASM) cell lines.

## Data Import

We need two things for this analysis:

- **countData**: a table with genes as rows and samples/experiments as columns 
- **colData**: metadata about the columns (i.e. samples) in the main countData objects

```{r import}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

Now lets have a peak at the data:

```{r}
head(metadata)
```

```{r}
head(counts)
```

## Check on metadata counts corespondance 

We need to check that the metadata matches the samlples in our count data. 

```{r}
ncol(counts) == nrow(metadata)
```

```{r}
colnames(counts)
metadata$id
```

> Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```


> Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control")
```

## Analysis Plan..

We have 4 replicates per conditions ("control" and "treated"). We want to compare the control vs the treated to see which genes expression levels change when we have the drug present. 

We can go down each gene and see if the average values in control column is different than the average treated column. 

1. Find which columns in `counts()` correspond to "control" samples
2. Extract/select these columns
3. Calculate an average value for each gene (i.e each row)

```{r}
# The indices (aka positions) that are "control"
control.inds <- metadata$dex == "control"
```

```{r}
# Extract/select these "control" columns from counts
control.counts <- counts[,control.inds]
```

```{r}
# Calculate the mean for each gene (i.e row)
control.means <- rowMeans(control.counts)
```

>Q Do the same for the "treated" samples - Find the mean count value per gene

```{r}
# The indices (aka positions) that are "treated"
treated.inds <- metadata$dex == "treated"
```

```{r}
# Extract/select these "treated" columns from counts
treated.counts <- counts[,treated.inds]
```

```{r}
# Calculate the mean for each gene (i.e row)
treated.means <- rowMeans(treated.counts)
```

Lets put these two mean values into a new data.frame, `meancounts`, for easy book keeping.

```{r}
meancounts <- data.frame(control.means,
                         treated.means)
head(meancounts)
```

>Q. Make a ggplot of average counts vs treated

```{r}
library(ggplot2)
ggplot(meancounts) +
  aes(x = control.means, y = treated.means) +
  geom_point(alpha=0.3) 
```

This is screaming to be log transformed as it is so highly skewed...

```{r}
ggplot(meancounts) +
  aes(x = control.means, y = treated.means) +
  geom_point(alpha=0.3) +
  scale_x_log10() +
  scale_y_log10()
```

## Log2 units and fold change

If we considered "treated"/"control" counts we will get a number that tells us the change.

```{r}
# No change
log2(20/20)
```

```{r}
# A doubling in the treated vs control
log2(40/20)
```

```{r}
log2(10/20)
```

> Q. Add a new column called `log2fc` for log2 fold change of treated/control to our `meancounts` object.

```{r}
meancounts$log2fc <- 
  log2(meancounts$treated.mean/
  meancounts$control.mean)
head(meancounts)
```


## Remove zero count genes

Typically we would not consider zero count genes - as we have no data bout them and they should be excluded from further consideration. These lead to "funky" log2 fold change values 9e.g. divide by zero errors ect.)

## DESeq analysis

We are missing any measure of significance from the work we have so far. Let's do this properly with **DESeq2** package.

```{r, message=FALSE}
library(DESeq2)
```

The DESeq2 package, like many bioconductor packages, want its imput in a very specific way - a data structure setup with all the info it needs for the calculations. 

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                       colData = metadata,
                       design = ~dex)
```

The main function in this package is called `DESeq()` it will run the full analysis for us on our `dds` imput object:

```{r}
dds <- DESeq(dds)
```

Exrtract our results:
```{r}
res <- results(dds)
head(res)
```

## Volcano Plot

A useful summary figure of our results is often called a volcano plot. Its basically a plot of log2 fold change values vs Adjusted P-values  


>Q. Use ggplot to make a first version "volcano plot" of `log2FoldChange` vs `padj`

```{r}
ggplot(res) +
  aes(log2FoldChange,padj) +
  geom_point()
```

This is not very useful because the y-axis(P-value) is not really helpful - we want to focus on low P-values

```{r}
ggplot(res) +
  aes(log2FoldChange,-log(padj)) +
  geom_point() +
  geom_vline(xintercept = c(-2, 2), color="red") +
  geom_hline(yintercept = -log(0.05),col="red")
```

## Add some plot annotaion

> Q. Add color to the points (genes) we care about, nice axis labels, and a useful title and a nice theme. 

```{r}
mycols <- rep("gray", nrow(res))
mycols[res$log2FoldChange >2] <- "blue"
mycols[res$log2FoldChange < -2] <- "darkgreen"
mycols[res$padj >= 0.05] <- "gray"
```


```{r}
ggplot(res) +
  aes(log2FoldChange,-log(padj)) +
  geom_point(col=mycols) +
  geom_vline(xintercept = c(-2, 2), color="red") +
  geom_hline(yintercept = -log(0.05),col="red")
```


## Save our results to a CSV file

```{r}
write.csv(res,file="results.csv")
```

