---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Paul Brencick (A17668863)"
format: html
toc: True
---

## Background

Today we will perform an RNA seq analysis on the effects of Dexamethasone (hereafter "Dex"), a common steroid, on airway smooth muscles (ASM) cell lines.

## Data Import

We need two things for this analysis:

- **countData**: a table with genes as rows and samples/experiments as columns 
- **colData**: metadata about the columns (i.e. samples) in the main countData objects

```{r import}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

Now lets have a peak at the data:

```{r}
head(metadata)
```

```{r}
head(counts)
```

## Check on metadata counts corespondance 

We need to check that the metadata matches the samlples in our count data. 

```{r}
ncol(counts) == nrow(metadata)
```

```{r}
colnames(counts)
metadata$id
```

> Q1. How many genes are in this dataset? 

```{r}
nrow(counts)
```


> Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control")
```

## Analysis Plan..

We have 4 replicates per conditions ("control" and "treated"). We want to compare the control vs the treated to see which genes expression levels change when we have the drug present. 

We can go down each gene and see if the average values in control collumn is different than the average treated collumn. 

1. Find which columns in `counts()` correspond to "control" samples
2. 


